---
title: "data_manipulation"
output: github_document
date: "2025-09-18"
---

```{r}
library(tidyverse)
```

Let's import datasets.

```{r}
litters_df=
  read_csv("dataset/FAS_litters.csv", na = c("NA", "", "."))
litters_df = 
    janitor::clean_names(litters_df)
pups_df = 
    read_csv("dataset/FAS_pups.csv",
        na = c(".", "NA", ""), 
        skip = 3)
pups_df = 
    janitor::clean_names(pups_df)
```

## `Select`

`select` selects variables from my dataframe.

select specific variables
```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range
```{r}
select(litters_df, group:gd_of_birth) #keeps variables from group to gd of birth
```

select by deletion
```{r}
select(litters_df, -group)
```

select by prefix - keeping everything that starts with gd in this case
```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange byb selecting
```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())
```

rename by selecting
```{r}
select(litters_df, )
```

what is i only select one variable - group in this case
```{r}
select(litters_df, group)

pull(litters_df, group) # you end up getting a vector here. 
                        #don't use a pull function!!!
```

you could assign the selection to object
```{r}
select(litters_df, litter_number, everything())

```

LA 1 code
```{r - Learning Assessment 1}
select(pups_df, litter_number, sex, pd_ears)
```

## filter

use `filter` to filter out rows

Here, get rid of everything that is not equal to 20 and rid of everything where pups born alive is not euqal to 6.
```{r}
filter(litters_df, gd_of_birth == 20, pups_born_alive == 6)
```

look for more than two pups alive
```{r}
filter(litters_df, pups_born_alive >= 5)
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive < 5)
filter(litters_df, pups_born_alive != 5) # not equal to 5

```

match to a set
```{r}
filter(litters_df, group %in% c("Con7", "Con8")) #selecting rows based on character variable
```

filter missing rows with `drop_na()`

```{r}
drop_na(litters_df) #drops all rows with NA in any columns
drop_na(litters_df, gd0_weight) #drops all rows with NA in gd0_weight
```

Learnign assessment 2
```{r}
filter(pups_df, sex == 1)
filter(pups_df, pd_walk< 11, sex ==2)
```

