---
title: "Data Import"
output: github_document
---

```{r}
library(tidyverse) #readr is part of tidyverse
library(readxl)
library(haven)
```

Lets import a dataset.

```{r}
litters_df =
  read_csv("data/FAS_litters.csv") # read_csv converts it into tibble
```

Look at your data!

```{r}
names(litters_df)
```

Variable names with spaces are a hassle. Let's change it to have no spaces.

```{r}
litters_df = janitor::clean_names(litters_df) #overwrite variable names using clean names function
names(litters_df)
```

Look at the data.

```{r}
litters_df # prints the data and shows what type of variables
```


Sometimes skimming data is neat?
```{r}
skimr::skim(litters_df) # provides high level summary
```

## Fix the missingness

We see in the data that there are some NA and some missing are indicated with "."

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))
```


## Import pups data

Here we account for the fact that first 3 rows are empty using skip option

```{r}
pups_df = 
    read_csv("data/FAS_pups.csv",
        na = c(".", "NA", ""), 
        skip = 3)

pups_df = 
    janitor::clean_names(pups_df)

skimr::skim(pups_df)
```

## Excel formats

CSVs are great but sometimes you get an excel file. 

```{r}
mlb_df = 
  read_excel("data/mlb11.xlsx")
```

Import LotR word counts, using specific range option

```{r}
fotr_df=
    read_excel("data/LOTR_Words.xlsx", range="B3:D6")
```

## SAS dataset import

Import the pulse dataset

```{r}
pulse_df = read_sas("data/public_pulse_data.sas7bdat")

pulse_df = 
  janitor::clean_names(pulse_df)

```

## Why do i hate read.csv so much

```{r eval = FALSE}
litters_df_base = 
  read.csv("data/FAS_litters.csv")
```

## what about data export

Use write_csv function.

```{r}
write_csv(fotr_df, "data/fotr_df.csv")

```

